---
title: "Circadian Mood"
output:
  word_document: default
  pdf_document: default
  html_document: default
date: '2022-11-21'
---

### Install missing packages
```{r}
pacman::p_load(tidyverse, lmerTest)

```


#Get and clean BodyFeelingExp data
```{r}
library(tidyverse)


path <- r"(C:\Users\maria\OneDrive\Dokumenter\Uni\Advanced Cognitive Neuroscience\homeostasis_and_happiness\BodyFeelingExp_data\BodyFeelingExp_data)"

setwd(path)

#Get FaceStroopData
datadir <- path


#Find files
files<-list.files(datadir,pattern='^BodyFeelingExp_.+?csv',full.names=TRUE)

#Prepare an empty data frame for the data (also removes old version)
dataBFE<-data.frame()

#How many datasets were there
n_datasets_raw<-length(files)
#Prepare a variable to monitor how many datasets we keep
n_datasets<-0
#Prepare a variable to monitor how many points we originally had
n_datapoints_raw<-0

#Loop to go through all files in the list
for(iii in 1:n_datasets_raw){
  
  #remove old loaded file to not risk importing it multiple times
  if(exists('data_temp')) rm(data_temp)
  
  #Load data
  data_temp<-read.csv(files[iii])
  if(dim(data_temp)[2]==31){
    data_temp[1,6]<-data_temp[dim(data_temp)[1],6]
    data_temp<-data_temp[1,c(6,8:27)]
        if(length(colnames(dataBFE))==0){
          dataBFE=data_temp
          rm(data_temp)
          #counter to monitor included datasets
          n_datasets<-n_datasets+1
        }
        #Bind loaded data with actual data
        else {dataBFE<-rbind(dataBFE,data_temp)
          rm(data_temp)
          #counter to monitor included datasets
          n_datasets<-n_datasets+1
        }
    }
}

#A variable to monitor how many points we keep
n_datapoints<-length(dataBFE[,1])

```

### Additional preprocessing
```{r}
#Make a variable which has hour and minutes of the day as decimal variable
dataBFE$hour2<-dataBFE$hour+(dataBFE$minute)/60
```

### summarizing stats -- assess IDs

```{r}
dataBFE %>%
  distinct(id) %>% 
  nrow()

dataBFE %>%
  group_by(id) %>%
  dplyr::summarise(count = length((id)))

# nrow(dataBFE)


```

## Take care of duplicate IDs

```{r}
# fix participant IDs
dataBFE <- dataBFE %>% 
  mutate(id = ifelse(id == " Woo55", "woo55", id)) %>% 
  mutate(id = ifelse(id == "Woo55", "woo55", id)) %>% 
  mutate(id = ifelse(id == "ESG44 ", "ESG44", id)) %>% 
  mutate(id = ifelse(id == " POP33", "POP33", id)) %>% 
  mutate(id = ifelse(id == "POP33 ", "POP33", id)) %>% 
  mutate(id = ifelse(id == "hej123", "hej12", id)) %>% 
  mutate(id = ifelse(id == "Qwel12", "qwel12", id)) %>% 
  mutate(id = ifelse(id == "www111", "www11", id)) %>% 
  mutate(id = ifelse(id == "Qwe12", "qwe12", id)) %>% 
  mutate(id = ifelse(id == "www123", "www12", id)) %>% 
  mutate(id = ifelse(id == "dig05", "dlg05", id))


```

## Some plotting

```{r}
library(ggplot2)
ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_point()
ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_point()
ggplot(dataBFE,aes(x=hour2,y=fresh))+geom_smooth()+geom_point()
ggplot(dataBFE,aes(x=hour2,y=hunger))+geom_smooth()+geom_point()

```
# High-school trigonometry for circidian analysis

## It all starts with a circle and a triangle
```{r}
#Radius
r<-1
# Get x-values from minus 1 to plus 1
x<-seq(-1,1,0.00001)
# A circle with c(0,0) centre can be written with these two equations (following Pythagoras)
y1<-sqrt(r^2-x^2)
y2<--sqrt(r^2-x^2)
y<-c(y1,y2)
x<-c(x,x)
#Plotting the circle with sine and cosine values
pp=pi/4
plot(x,y,type='l')
lines(x=c(0,0),y=c(-1,1))
lines(x=c(0,cos(pp)),y=c(0,sin(pp)),col='darkgreen')
lines(x=c(cos(pp),cos(pp)),y=c(0,sin(pp)),col='darkblue')
text(x=c(-0.15+cos(pp)),y=c(0.5*sin(pp)),labels='sin(x)',col='darkblue')
lines(x=c(0,cos(pp)),y=c(sin(pp),sin(pp)),col='darkred')
text(x=c(0.5*cos(pp)),y=c(+0.15+sin(pp)),labels='cos(x)',col='darkred')
```

### Going beyond the circle, the sine and cosine functions can describe cycles in time

$$
y(t)=\beta{sin}(2\pi f t )
$$
$$
y(t)=\beta{cos}(2\pi f t )
$$
Beta is the amplitude (the height of the wave)
f is the frequency (the number of cycles per time unit)
t is time point



#### Plotting circadian sine and cosine waves
```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/24
#sample frequency per hour
fs=100
#Duration in hours
dur=24
#A time vector divided by fs
t = seq(0, dur, 1/fs)
#Make a sine wave (with amplitude =1) for each time point in t
u = sin(2*pi*cf*t)
#Make a cosine wave (with amplitude =1) for each time point in t
u2= cos(2*pi*cf*t)
#Plot the waves
plot(x=t,y=u, type='l',col='darkblue',xlab='time(hours)',ylab='Amplitude')
text(x=1+t[1000],y=-0.2+u[500],labels='y=sin(2*pi*cf*t)',col='darkblue')
lines(x=t,y=u2, type='l',col='darkred')
text(x=-1+t[500],y=-0.2+u2[500],labels='Y=cos(2*pi*cf*t)',col='darkred')
```
As can be seen, the sine function is a time shifted version of the cosine function and vice versa.
The time shift is also called phase. We can add a constant for the phase (phi) to the sine/cosine wave function.

$$
y(t)=\beta{sin}(2\pi f t + \phi)
$$

#### Plotting circadian sine wave with phase shift
```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/24
#sample frequency per hour
fs=100
#Duration in hours
dur=24
#A time vector divided by fs
t = seq(0, dur, 1/fs)
#a  phase shift of pi/2 radians (half a cycle) This could be any number
phi=pi/2
#Make a sine wave (with amplitude =1)
u = sin(2*pi*cf*t)
#Make a sine wave (with amplitude =1), and phase shift
u2= cos(2*pi*cf*t+phi)
#Plot the waves
plot(x=t,y=u, type='l',col='darkblue',xlab='time(hours)',ylab='Amplitude')
text(x=1+t[1000],y=-0.2+u[500],labels='y=sin(2*pi*cf*t)',col='darkblue')
lines(x=t,y=u2, type='l',lty='dashed',col='darkblue')
text(x=-1+t[500],y=0.2+u2[500],labels='Y=sin(2*pi*cf*t+phi)',col='darkblue')
```
Using the trigonometric identity
^caption should be sine and cosine

$$
{sin(A+B)}={sin(A)cos(B)}-{cos(A)sin(B)}
$$
we can rewrite the sine function (including phase) as

$$
y(t)={\beta_1{sin}(2\pi f t)}+{\beta_2{cos}(2\pi f t)}
$$
^so we have a beta estimate for sin and one for cos
where 
$$
{\beta_1}=\beta{cos(\phi)}, {\beta_2}=-\beta{sin(\phi)}
$$
We can use the rewritten sine function in a linear regression analysis, where we estimate the best fitting B1 and B2. This will yield a composite estimate of the amplitude and the phase of the data.

This will allow us to use sine and cosine waves to model a circadian rhythm, even if we don't know when it peaks. The amplitude will be given by

$$
{\beta}=\sqrt{\beta_1^2+\beta_2^2}
$$
The phase (phi) will be given by
$$
{\phi}=atan2(\beta_1,\beta_2)
$$

#### Use the sine and cosine waves to make 24 hour oscillation predictors for the BodyFeelingExp data

```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/24

#Make sine and cosine waves for each time point present in the data

dataBFE$sinCirc<-sin(2*pi*cf*dataBFE$hour2)
dataBFE$cosCirc<-cos(2*pi*cf*dataBFE$hour2)

# Plot the predictors for each data point in the data
ggplot(dataBFE, aes(x=hour2,y=sinCirc))+
      geom_point(col='darkblue')+
      geom_point(aes(y=cosCirc),col='darkred')+
      ylab('Circadian sine and cosine')+xlab('Time of day (hours)')

```

## Fitting 24 hour Oscillation model

#### Freshness
```{r}
library(lmerTest)
# Freshness: Simple oscillation model
modelBFEfreshCirc<-lmer(fresh~sinCirc+cosCirc+(1|id),data=dataBFE)
m_temp<-summary(modelBFEfreshCirc)
m_temp
dataBFE$freshCircFit<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc

plot(x=dataBFE$hour2,y=dataBFE$freshCircFit,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=fresh))+geom_smooth()+geom_smooth(aes(x=hour2,y=freshCircFit),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')
```


#### Mood
```{r}
library(lmerTest)
# Mood: Simple oscillation model
modelBFEmoodCirc<-lmer(mood~sinCirc+cosCirc+(1|id),data=dataBFE)
m_temp<-summary(modelBFEmoodCirc)
m_temp
dataBFE$CircFit<-m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc

plot(x=dataBFE$hour2,y=dataBFE$CircFit,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_smooth(aes(x=hour2,y=CircFit+mean(mood)),col='darkgreen')+labs(x='Time of day (hours)', y='Negative/Positive Mood')
```

#### HeartRate
```{r}
# Freshness: Simple oscillation model
modelBFEHeartRateCirc<-lmer(HeartRate~sinCirc+cosCirc+(1|id),data=dataBFE)
m_temp<-summary(modelBFEHeartRateCirc)
m_temp
dataBFE$HeartRateCircFit<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc

plot(x=dataBFE$hour2,y=dataBFE$HeartRateCircFit,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_smooth(aes(x=hour2,y=HeartRateCircFit),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')
```


#### Make a 12 hour oscillation model to compare with the 24 hour model

```{r}
#cycle frequency (in this case per hour) - one cycle per 24 hours
cf=1/12

#Make sine and cosine waves for each time point present in the data

dataBFE$sinCirc12<-sin(2*pi*cf*dataBFE$hour2)
dataBFE$cosCirc12<-cos(2*pi*cf*dataBFE$hour2)

# Plot the predictors for each data point in the data
ggplot(dataBFE, aes(x=hour2,y=sinCirc12))+
      geom_point(col='darkblue')+
      geom_point(aes(y=cosCirc12),col='darkred')+
      ylab('sinCirc/cosCirc')+
      ggtitle('Sine and cosine waves from time point in the data')+
      xlab('Hours in a day')+
      ylab('Circadian sine and cosine')

```

## Fitting 12 hour Oscillation models


#### Freshness
```{r}
# Freshness: Simple oscillation model
modelBFEfreshCirc12<-lmer(fresh~sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEfreshCirc12)
m_temp
dataBFE$freshCircFit12<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc12+m_temp$coefficients[3,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$freshCircFit12,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=fresh))+geom_smooth()+geom_smooth(aes(x=hour2,y=freshCircFit12),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')
```


#### Mood
```{r}
library(lmerTest)
# Mood: Simple oscillation model
modelBFEmoodCirc12<-lmer(mood~sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEmoodCirc12)
m_temp
dataBFE$CircFit12<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc12+m_temp$coefficients[3,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$CircFit12,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_smooth(aes(x=hour2,y=CircFit12),col='darkgreen')+labs(x='Time of day (hours)', y='Negative/Positive Mood')
```

#### HeartRate
```{r}
# Freshness: Simple oscillation model
modelBFEHeartRateCirc12<-lmer(HeartRate~sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEHeartRateCirc12)
m_temp
dataBFE$HeartRateCircFit12<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc12+m_temp$coefficients[3,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$HeartRateCircFit12,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_smooth(aes(x=hour2,y=HeartRateCircFit12),col='darkgreen')+labs(x='Time of day (hours)', y='Tired/Fresh')
```
## Combining 12 and 24 hours

#### Freshness
```{r}
# Freshness: Simple oscillation model
modelBFEfreshCircBoth<-lmer(fresh~sinCirc+cosCirc+sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEfreshCircBoth)
m_temp
dataBFE$freshCircFitBoth<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc+m_temp$coefficients[4,1]*dataBFE$sinCirc12+m_temp$coefficients[5,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$freshCircFitBoth,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=fresh))+
  geom_smooth()+
  geom_smooth(aes(x=hour2,y=freshCircFitBoth),col='darkgreen')+
  labs(x='Time of day (hours)', y='Alertness')

  
```

#### Mood
```{r}
library(lmerTest)
# Mood: Simple oscillation model
modelBFEmoodCircBoth<-lmer(mood~sinCirc+cosCirc+sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEmoodCircBoth)
m_temp
dataBFE$CircFitBoth<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc+m_temp$coefficients[4,1]*dataBFE$sinCirc12+m_temp$coefficients[5,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$CircFitBoth,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=mood))+geom_smooth()+geom_smooth(aes(x=hour2,y=CircFitBoth),col='darkgreen')+labs(x='Time of day (hours)', y='Mood')
```
#### HeartRate
```{r}
# Freshness: Simple oscillation model
modelBFEHeartRateCircBoth<-lmer(HeartRate~+sinCirc+cosCirc+sinCirc12+cosCirc12+(1|id),data=dataBFE)
m_temp<-summary(modelBFEHeartRateCircBoth)
m_temp
dataBFE$HeartRateCircFitBoth<-m_temp$coefficients[1,1]+m_temp$coefficients[2,1]*dataBFE$sinCirc+m_temp$coefficients[3,1]*dataBFE$cosCirc++m_temp$coefficients[4,1]*dataBFE$sinCirc12+m_temp$coefficients[5,1]*dataBFE$cosCirc12

plot(x=dataBFE$hour2,y=dataBFE$HeartRateCircFitBoth,type='p',col='darkgreen')

ggplot(dataBFE,aes(x=hour2,y=HeartRate))+geom_smooth()+geom_smooth(aes(x=hour2,y=HeartRateCircFitBoth),col='darkgreen')+labs(x='Time of day (hours)', y='Heartrate')
```

## Model comparisons
### Is the 24 hour or the 12 hour model the best?
```{r}
#Comparing models for freshness
anova(modelBFEfreshCirc12,  modelBFEfreshCircBoth)
anova( modelBFEfreshCirc, modelBFEfreshCircBoth)

#Comparing models for mood
anova(modelBFEmoodCirc12, modelBFEmoodCircBoth)
anova(modelBFEmoodCirc, modelBFEmoodCircBoth)

#Comparing models for heartrate
anova(modelBFEHeartRateCirc12,modelBFEHeartRateCircBoth)
anova(modelBFEHeartRateCirc,modelBFEHeartRateCircBoth)

#effect(modelBFEHeartRateCirc)
```
```{r}


```




### Activity and Mental activity
```{r}

library(ggplot2)
ggplot(dataBFE,aes(x=MentalActivity))+geom_bar()
ggplot(dataBFE,aes(x=Activity))+geom_bar()+coord_flip()

#Mood as function of Mental activity
modelBFEmoodMentalAct<-lmer(mood~MentalActivity + (1|id),data=dataBFE)
anova(modelBFEmoodMentalAct)

#Mood as function of activity
modelBFEmoodAct<-lmer(mood~Activity + (1|id),data=dataBFE)
anova(modelBFEmoodAct)
```
### alertness
```{r}
# alertness boxplot
ggplot(dataBFE, aes(as.factor(month), fresh, fill = id)) + 
  geom_boxplot() + labs(x = "Month", y = "Alertness")+
  ggtitle("Alertness per month per participant")
```

### heartrate
```{r}
# heartrate boxplot
ggplot(dataBFE, aes(as.factor(month), HeartRate, fill = id)) + 
  geom_boxplot() + labs(x = "Month", y = "Heartrate")+
  ggtitle("Heartrate per month per participant")
```
```{r}
# mood scatterplot circadian sine
ggplot(dataBFE, aes(sinCirc, mood, colour = id)) + 
  # geom_smooth(method = "loess") +
  geom_point() + labs(x = "Participants", y = "Mood") +
  ggtitle("Mood and circadian sine wave")
```


```{r}
# mood by activity
ggplot(dataBFE, aes(mood, Activity, colour = Activity)) + 
  # geom_point()+
  geom_boxplot()+
  ggtitle("Activity and mood")+xlab('Mood')
```

  
### Mood
```{r}
# mood boxplot
ggplot(dataBFE, aes(as.factor(month), mood, fill = id)) + 
  geom_boxplot() + labs(x = "Month", y = "Mood") +
  ggtitle("Mood per month per participant")
```
```{r}
# mood scatterplot
ggplot(dataBFE, aes(as.factor(id), mood, colour = id)) + 
  geom_point() + labs(x = "Participants", y = "Mood") +
  ggtitle("Mood variability per participant")

```


### Mood by month
```{r}
library(ggplot2)
ggplot(dataBFE,aes(y=mood, x=as.factor(month)))+geom_boxplot()


```